\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-07 k-planar.tex
Cleaning up the 1-planar proof

Let $G$ be an edge-maximal 1-plane multigraph with no two parallel edges on the boundary of a single face.  Here, edge-maximal should be taken to mean that, if any two vertices $v$ and $w$ appear on a common face\footnote{The \defin{faces} of an embedded graph $G$ are the connected components of $\R^2\setminus \bigcup_{vw\in E(G)} vw$.  We say that a vertex $v\in V(G)$ appears on a face $F$ if $v$ is contained in the closure of $F$.} $F$, then there is an edge $vw\in E(G)$ that is contained in the boundary of $F$.  \note{DW}{Each face of $G$ is one of two types, either bounded by three edges (none of which are crossed), or bounded by one edge that is not crossed plus portions of two crossing edges. Right? If this is true, we should say it here, which would help to clear up the confusion below. Add a figure.}
We assume that no two edges incident to a common vertex cross each other since, in a 1-plane graph, such a crossing can always be removed by a local modification to obtain an isomorphic 1-plane graph in which the two edges do not cross.\footnote{While this is true for 1-plane graphs it is not true for $k$-plane graphs with $k\ge 3$; the uncrossing operation can increase the number of crossings on a particular edge from $k$ to $2(k-1)$.}

\note{DW}{I suggest we say that it follows from Euler's formula that $G$ is finite. (To see this, say $G$ has $n$ vertices, $m$ non-crossed edges, and $k$ crossings. Let $G'$ be the plane graph obtained from $G$ by adding a dummy vertex at each crossing. So $G'$ is a plane multigraph in which every face is bounded by a 3-cycle. It follows from Euler's formula that $m+4k = |E(G')| = 3( |V(G')|-2) = 3 ( n+k-2)$, implying	$m+k = 3 (n-2)$ and $|E(G)| = m+2k \leq 6(n-2)$, which can also be shown by deleting one edge from each crossing point. ) }

\note{DW}{It seems to be that $G$ can be obtained from a plane multigraph with faces of size 3 or 4 such that no edge is in two 3-faces, by adding a pair of crossing edges across each 4-face. I think this characterises edge-maximal 1-planar multigraphs with no 2-faces. Right? Does this viewpoint help? Is it known?}

\note{DW}{Say $G$ is obtained from a plane multigraph by making each face a clique. Is $G\subseteq H \boxtimes P \boxtimes K_{f(\omega(G)}$? for some treewidth 3 graph $H$.}

A \defin{kite} in $G$ is the subgraph $K=G[\{v,w,x,y\}]$ induced by the endpoints of a pair of crossing edges $vw,xy\in E(G)$.  It follows from edge-maximality that every kite is isomorphic to the complete graph $K_4$. \note{DW}{We need to define a kite to be the subgraph with vertex-set $\{v,w,x,y\}$ and edge-set $\{vx,xy,vx,vy,wy,wx\}$ where $vx,vy,wy,wx$ are the edges on the `outside' of the kite-faces. Then it really is a simple $K_4$. I think this is what we intended. We should add a figure showing a kite and kite-faces. }
The edges $vw$ and $xy$ are called \defin{spars} of $K$.  The cycle $vxwy$ is called the \defin{sail} of $K$.  It follows from edge-maximality that none of the edges $vx$, $xw$, $wy$, or $yv$ are crossed by any other edges of $G$.

\referee{2}{Page 13. Perhaps I am misunderstanding something, but I do not see why
'none of the edges $vx$, $xw$, $wy$, or $yv$ are crossed by any other edges of
$G$.'  For example, see the attached PDF for a picture where $vw$, $xw$, $wy$,
and $yv$ are all crossed by other edges of $G$.\\
\includegraphics[width=60mm]{kite}}
\note{DW}{We need to explain more about how edge-maximality fixes this problem. In this case we can draw other $xv$, $vy$, $yw$ $wx$ edges close to the central crossing edges.} Thus any edge that is a spar of a kite $K$ is not part of a sail of any kite $K'$.

\referee{2}{Page 13. The attached PDF also shows that the definition of 'kite
face' may not be well-defined.  For example, I do not see why there
cannot be some edges of $G$ 'inside' a kite face.  Even if this is not
the case, I think it is more precise to say that a kite face has 'two
and a half edges' and two vertices of $G$ on its boundary' rather than
'three edges and two vertices of $G$ on its boundary'.}
\note{DW}{This should be fixed if we define kite more precisely. A kite-face is bounded by one edge and two half-edges right?}

\referee{2}{Page 13. Could it be that the subgraph induced by the vertices of a
	kite is a $K_4$ with some parallel edges?  This seems relevant later in
	the proof.}

\note{DW}{If we define kite as above, this is not a problem}

The 1-plane graph $G$ has a plane triangulation $G'$ as a subgraph that can be obtained by removing one spar from each kite in $G$.  Observe that, for any spar $xy\in E(G)\setminus E(G')$ that crosses $vw\in E(G')$, $G'$ contains the path $vxw$ (and $vyw$).  It follows that $\dist_{G'}(v,w)\le 2$.

>>>>

Let $G$ be an edge-maximal $1$-plane multigraph with no two parallel edges on the boundary of a single face.  Here, edge-maximal should be taken to mean that, if any two vertices $v$ and $w$ appear on a common face\footnote{The \defin{faces} of an embedded graph $G$ are the connected components of $\R^2\setminus \bigcup_{vw\in E(G)} vw$.  We say that a vertex $v\in V(G)$ appears on a face $F$ if $v$ is contained in the closure of $F$.} $F$, then there is an edge $vw\in E(G)$ that is contained in the boundary of $F$.  Refer to \cref{edge_maximal}. We assume that no two edges incident to a common vertex of $G$ cross each other since, in a 1-plane graph, such a crossing can always be removed by a local modification to obtain an isomorphic 1-plane graph in which the two edges do not cross.\footnote{While this is true for 1-plane graphs it is not true for $k$-plane graphs with $k\ge 3$; the uncrossing operation can increase the number of crossings on a particular edge from $k$ to $2(k-1)$.}

\begin{figure}
  \begin{center}
    \begin{tabular}{c@{\hspace{1cm}}c}
      \includegraphics{figs/one_planar_example-1} &
      \includegraphics{figs/one_planar_example-2} \\
      (a) & (b)
    \end{tabular}
  \end{center}
  \caption{Examples of (a)~a $1$-plane graph that is not edge maximal; and (b)~an edge-maximal $1$-plane graph.}
\end{figure}

To understand the structure of an edge maximal $1$-plane multigraph $G$, it is helpful to consider the plane graph $G_0$ obtained by replacing each pair of crossing edges $vw$ and $xy$ with four edges $vp$, $wp$, $xp$, and $yp$ meeting at a newly added \emph{dummy vertex} $p$. Let $F$ be a face of $G_0$ and let $W:=w_0,\ldots,w_r$ be the facial walk around $F$. By edge maximality, if $W$ contains only non-dummy vertices, then it contains exactly three vertices and $F$ is bounded by three edges of $G$.

Furthermore, for any edge $w_iw_{i+1}$ of $W$, at least one of $w_i$ or $w_{i+1}$ is not a dummy vertex since, otherwise this would imply that the edge of $G$ that contains $w_iw_{i+1}$ is involved in at least two crossings.  Therefore each dummy vertex $w_i$ of $W$ is surrounded by two vertices $w_{i-1},w_{i+1}\in V(G)$.  Since $w_i$ has degree $4$ in $G_0$, $w_{i-1}\neq w_i$.  Since $G$ is edge-maximal the edge $w_{i-1}w_i$ is an edge of $G$ on the boundary of $F$.  Therefore, if $F$ has a dummy vertex in its facial walk $W$, then $W=w_{i-1},w_i,w_{i+1}$ and $F$ is bounded by three edges of $G_0$, each of which is contained in a different edge of $G$.

Therefore, in such a graph each face of $G$ is one of two types:
\begin{inparaenum}[(i)] 
  \item bounded by three edges (none of which are crossed) or 
  \item bounded by one edge that is not crossed plus portions of two edges that cross each other.
\end{inparaenum}
Consider some crossing $(p,vw,xy)$ in $G$.  The point $p$ at which this crossing occurs is a vertex of $G_0$ that has degree $4$.  There are four faces $F_1,\ldots,F_4$ of $G_0$ that contain $p$.  Each of these four faces of is of Type~(ii), so the boundary of $F_1\cup\cdots\cup F_4$ includes a $4$-cycle $C:=vxwy$ in $G$ whose edges are not involved in any crossings.  We call the subgraph $K:=C\cup\{vw,xy\}$ a \defin{kite} of $G$.  The edges $vw$ and $xy$ are called \defin{spars} of $K$.  The cycle $vxwy$ is called the \defin{sail} of $K$.  Thus any edge that is a spar of a kite $K$ is not part of a sail of any kite $K'$.

\note{PM}{I've commented out a lot of red and blue here to make this readable.}
% 
% It follows from edge-maximality that none of the edges $vx$, $xw$, $wy$, or $yv$ are crossed by any other edges of $G$.

% 
% \note{DW}{Each face of $G$ is one of two types, either bounded by three edges (none of which are crossed), or bounded by one edge that is not crossed plus portions of two crossing edges. Right? If this is true, we should say it here, which would help to clear up the confusion below. Add a figure.}
% 
% \note{DW}{I suggest we say that it follows from Euler's formula that $G$ is finite. (To see this, say $G$ has $n$ vertices, $m$ non-crossed edges, and $k$ crossings. Let $G'$ be the plane graph obtained from $G$ by adding a dummy vertex at each crossing. So $G'$ is a plane multigraph in which every face is bounded by a 3-cycle. It follows from Euler's formula that $m+4k = |E(G')| = 3( |V(G')|-2) = 3 ( n+k-2)$, implying	$m+k = 3 (n-2)$ and $|E(G)| = m+2k \leq 6(n-2)$, which can also be shown by deleting one edge from each crossing point. ) }
% 
% \note{DW}{It seems to be that $G$ can be obtained from a plane multigraph with faces of size 3 or 4 such that no edge is in two 3-faces, by adding a pair of crossing edges across each 4-face. I think this characterises edge-maximal 1-planar multigraphs with no 2-faces. Right? Does this viewpoint help? Is it known?}
% 
% \note{DW}{Say $G$ is obtained from a plane multigraph by making each face a clique. Is $G\subseteq H \boxtimes P \boxtimes K_{f(\omega(G)}$? for some treewidth 3 graph $H$.}
% 
% % A \defin{kite} in $G$ is the subgraph $K=G[\{v,w,x,y\}]$ induced by the endpoints of a pair of crossing edges $vw,xy\in E(G)$.  It follows from edge-maximality that every kite is isomorphic to the complete graph $K_4$. \note{DW}{We need to define a kite to be the subgraph with vertex-set $\{v,w,x,y\}$ and edge-set $\{vx,xy,vx,vy,wy,wx\}$ where $vx,vy,wy,wx$ are the edges on the `outside' of the kite-faces. Then it really is a simple $K_4$. I think this is what we intended. We should add a figure showing a kite and kite-faces. }
% 
% \referee{2}{Page 13. Perhaps I am misunderstanding something, but I do not see why
% 'none of the edges $vx$, $xw$, $wy$, or $yv$ are crossed by any other edges of
% $G$.'  For example, see the attached PDF for a picture where $vw$, $xw$, $wy$,
% and $yv$ are all crossed by other edges of $G$.\\
% \includegraphics[width=60mm]{kite}}
% \note{DW}{We need to explain more about how edge-maximality fixes this problem. In this case we can draw other $xv$, $vy$, $yw$ $wx$ edges close to the central crossing edges.} \note{PM}{In the example above, $v$ and $y$ appear on a common (inner) face $F$, so by definition of edge-maximality there is an edge $xy$  on the boundary of $F$. }
% 
% \referee{2}{Page 13. The attached PDF also shows that the definition of 'kite
% face' may not be well-defined.  For example, I do not see why there
% cannot be some edges of $G$ 'inside' a kite face.  Even if this is not
% the case, I think it is more precise to say that a kite face has 'two
% and a half edges' and two vertices of $G$ on its boundary' rather than
% 'three edges and two vertices of $G$ on its boundary'.}
% \note{DW}{This should be fixed if we define kite more precisely. A kite-face is bounded by one edge and two half-edges right?}
% 
% \referee{2}{Page 13. Could it be that the subgraph induced by the vertices of a
% 	kite is a $K_4$ with some parallel edges?  This seems relevant later in
% 	the proof.}
% 
% \note{DW}{If we define kite as above, this is not a problem}

By removing one spar from each kite of $G$ we obtain a plane multigraph $G'$ each of whose faces is bounded by three edges.  Observe that, for any spar $xy\in E(G)\setminus E(G')$ that crosses $vw\in E(G')$, $G'$ contains the path $vxw$ (and $vyw$).  It follows that $\dist_{G'}(v,w)\le 2$.
